---
title: "Gapminder Data Exploration"
output: 
    html_document:
        theme: cerulean
        toc: true
        keep_md: true

---
## Setting up access to the dataset

List all the available datasets in R
```{r}
data(package = .packages(all.available = TRUE))
```

I will be analyzing gapminder dataset. First, install the `gapminder` R package using this command: install.packages('gapminder')

If you already have the package installed, access the package

```{r}
library(gapminder)
```

## Exploring of data frames

Previewing dataset with the head function
```{r}
head(gapminder)
```

Explore some of its properties

Provide a summary of variables and its quantitative characteristics
```{r}
summary(gapminder)
```

Display the internal structure to the dataset (an alternative to summary)
```{r}
str(gapminder)
```

Measure the number of rows
```{r}
nrow(gapminder)
```

Measure the number of columns
```{r}
ncol(gapminder)
```

Alternatively, you can find dataset's dimensions
```{r}
dim(gapminder)
```

## Exploring Variables

The dataset consists of panel data: 3 variables (life expectancy, population and GDP per capita) are recorded from 1952 to 2007 for each country and the corresponding continent (some data is missing). 

### Individual statistics
We can measure various summary statistics (means, medians, max/min, etc.) for our variables by adding the variable name after $.

Find mean, max, min, standard deviation, and variance for the life expectancy variable
```{r}
mean(gapminder$lifeExp)
max(gapminder$lifeExp)
min(gapminder$lifeExp)
sd(gapminder$lifeExp)
var(gapminder$lifeExp)
```

### Adding If statements

If you wanted to know whether the GDP per capita for any country in any given year has ever exceeded 110,000 USD, you can add the following if statement
```{r}
if (max(gapminder$gdpPercap) > 110000){
  print('The GDP per capita for one of the countries has exceeded 110,000 USD.')
}else{
  print('The GDP per capita for any of the countries has never exceeded 110,000 USD.')
}
```

### Printing desired outcomes
You can print any of the selected summary statistics

```{r}
print('The standard deviation of the life expectancy variable is')
print(sd(gapminder$lifeExp))
```

### Group statistics
We can find various summary statistics (means, medians, max/min, etc.) for different groups using the 'aggregate' function.

Find the average population size of a country in a given year
```{r}
aggregate(pop ~ year, gapminder, mean)
```

Find max GDP per capita by continent
```{r}
aggregate(gdpPercap ~ continent, gapminder, max)
```

Find median life expectancy for each country across all years
```{r}
aggregate(lifeExp ~ country, gapminder, median)
```


### Adding selection condtions

Download dplyr package to help us sort, summarize and filter tabular data using this command: install.packages('dplyr')

If the package is already installed, get access to it
```{r}
library(dplyr)
```

I have an option to only work with the selected subgroups. For instance, I want to find an average population of a Europen country in 2005. Notice how you can add multiple conditions here.
```{r}
    filter(gapminder, continent == "Europe", year == 2005)
    MeanPopulation <- mean(gapminder$pop)
    print(MeanPopulation)
```

## Visualizing data


### Simple plots
We can visualize the relationships between these variables. 

Here is a scatterplot between population size (on the x-axis) and GDP per capita (on the y-axis)
```{r}
plot(gapminder$pop, gapminder$gdpPercap)
```

Here is a scatterplot between time (measured in years on the x-axis) and GDP per capita (on the y-axis)
```{r}
plot(gapminder$year, gapminder$gdpPercap)
```

### Complex plots

If we want to be able to capture more complexity in our groups, we need to install the ggplot package using this command: install.packages('ggplot2')

If the package is already installed, get access to it
```{r}
library(ggplot2)
```

Here we can visualize various continents by using colors.
```{r}
ggplot(gapminder, aes(x=pop, y=gdpPercap, color=continent))+geom_point()+ xlab("Country population size") + 
  ylab("GDP per capita") 
```
